---
title: "Experiment"
author: "Richel J.C. Bilderbeek"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Experiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette shows how to use `pirouette` in a scientific experiment.

Remember the main question `pirouette` helps to answer:

> Would we know the true phylogeny in nature,
> how well would we be able to infer it 
> from a DNA alignment?

In this vignette, we investigate the effect of extintion rate in inferring
a phylogeny from a (constant rate) birth-death process. We expect that
inference is better if there is more information to work on. For example,
when there are more taxa. But extinctions do not only decrease the number
of taxa, also the information when these extinct taxa cam into exinstence
is lost. As extinctions lower the amount of information, we predict
a higher inference error for higher extinction rates.

Our experimental setup will not be the best setup. A good experimental
setup will have, among others, more replicates, longer MCMC runs
and a more equal comparison.

## Setup

We specify the extinction rates we will investigate:

```{r}
ext_rates <- seq(0.0, 0.4, length.out = 3)
```

For a speciation rate, we use one value:

```{r}
spec_rate <- 1.0
```

Now, per extinction rate, we simulate one phylogeny:

```{r fig.width=7, fig.height=7}
set.seed(42)
crown_age <- 4.0
phylogenies <- list()
for (i in seq_along(ext_rates)) {
  ext_rate <- ext_rates[i]
  phylogeny <- becosys::bd_sim(
    bd_params = becosys::create_bd_params(
      spec_rate = spec_rate,  
      ext_rate = ext_rate
    ),
    crown_age = crown_age,
    conditioned_on = "non_extinction"
  )$reconstructed_tree
  phylogenies[[i]] <- phylogeny
}
ape::plot.phylo(phylogenies[[1]])
ape::plot.phylo(phylogenies[[2]])
ape::plot.phylo(phylogenies[[3]])
```


Now, per phylogeny, we estimate the error BEAST2 makes:

```{r}

alignment_params <- pirouette::create_alignment_params(
  root_sequence = "aaaaccccggggtttt",
  mutation_rate = create_standard_mutation_rate(phylogeny)
)
model_select_param <- pirouette::create_gen_model_select_param(
  alignment_params = alignment_params,
  tree_prior = beautier::create_bd_tree_prior()
)
inference_params <- pirouette::create_inference_params(
  mcmc = beautier::create_mcmc(chain_length = 5000, store_every = 1000)
)
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  model_select_params = model_select_params,
  inference_params = inference_params
)
errors <- list()
if (rappdirs::app_dir()$os != "win") {
  for (i in seq_along(ext_rates)) {
    phylogeny <- phylogenies[[i]]
    df <- pirouette::pir_run(
      phylogeny = phylogeny,
      pir_params = pir_params
    )
    errors[[i]] <- df
  }
}
```

Now, we put all data in one data frame:

```{r}
if (rappdirs::app_dir()$os != "win") {
  first_error_col <- which(colnames(errors[[1]]) == "error_1")
  last_error_col <- ncol(errors[[1]])
  n_errors <- 1 + last_error_col - first_error_col
  df <- data.frame(ext_rate = as.factor(rep(ext_rates, each = n_errors)), idx = seq(1, n_errors), error = NA)
  
  for (i in seq_along(ext_rates)) {
    this_df <- errors[[i]]
    nltts <- this_df[1, first_error_col:last_error_col]
    to_row_index <- 1 + (i * n_errors) - n_errors 
    df$error[to_row_index:(to_row_index + n_errors - 1)] <- t(as.numeric(nltts)) 
  }
}
```

Plotting it:

```{r fig.width=7, fig.height=7}
if (rappdirs::app_dir()$os != "win") {
  ggplot2::ggplot(
    df, ggplot2::aes(x = ext_rate, y = error)
  ) + ggplot2::geom_boxplot()
  ggplot2::ggplot(
    df, ggplot2::aes(x = ext_rate, y = error)
  ) + ggplot2::geom_violin()
}
```
