---
title: "Partial function evaluation in pirouette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Partial function evaluation in pirouette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# knitr::opts_chunk$set(
#   collapse = TRUE,
#   comment = "#>"
# )
# Include this library ahead of time, as it gives
# messages upon lazy loading
library(geiger)
```

This vignette describes partial function evaluation used in `pirouette`.

```{r, message = FALSE}
library(pirouette)
```

We'll use `pryr` (from
the Tidyverse to get that functionality.

```{r, message = FALSE}
library(pryr)
```

## The functions

These are the functions we'll create, in order of increasing complexity:

 1. Create a twin tree
 2. Create a true alignment
 3. Create a twin alignment


## 1. create a twin tree

We start from a simple true tree:

```{r}
true_phylogeny <- ape::read.tree(text = "((A:1, B:1):1, C:2);")
ape::plot.phylo(true_phylogeny)
```

We want to create a twin tree from it. There is a function that does
so:

```{r}
twin_tree <- pirouette::create_twin_tree(true_phylogeny)
ape::plot.phylo(twin_tree)
```

The `create_twin_tree`, however, creates the twin tree in the way specified
in its other function argument: the `twinning_params`.

```{r}
twinning_params <- create_twinning_params()
```

Calling `create_twin_tree` results in the same twin tree:

```{r}
twin_tree <- pirouette::create_twin_tree(
  phylogeny = true_phylogeny,
  twinning_params = twinning_params
)
ape::plot.phylo(twin_tree)
```

How to specify *how* the twin tree is created? Use 
the `twinning_params$sim_twin_tree_function`:

```{r}
twinning_params <- create_twinning_params(
  sim_twin_tree_function = create_sim_bd_twin_tree_function()
)
twin_tree <- pirouette::create_twin_tree(
  phylogeny = true_phylogeny,
  twinning_params = twinning_params
)
ape::plot.phylo(twin_tree)
```

Due to this architecture, `twinning_params$sim_twin_tree_function` must
be a function that takes *one* argument, which is the true phylogeny.

Here is an example that creates a twin tree by simulation 
a random coalescent tree with the same number of tips:

```{r}
# Define my function
my_function = function(true_phylogeny) {
  ape::rcoal(n = ape::Ntip(true_phylogeny))
}
# Put my function in the twinning_params
twinning_params <- create_twinning_params(
  sim_twin_tree_function = my_function
)
# Create a twin tree using my function
twin_tree <- pirouette::create_twin_tree(
  phylogeny = true_phylogeny,
  twinning_params = twinning_params
)
# Show the twin tree created by my function
ape::plot.phylo(twin_tree)
```

So, `twinning_params$sim_twin_tree_function` must
be a function that takes *one* argument, which is the true phylogeny. What
if we want a function that indeed has one argument for the true phylogeny,
but also some additional ones?

An example is `sim_bd_twin_tree`, which uses a seed, 
a method and a number of replicates:

```{r}
head(sim_bd_twin_tree)
```

We cannot simply plug it in:

```{r}
tryCatch(
  create_twinning_params(
    sim_twin_tree_function = sim_bd_twin_tree
  ),
  error = function(e) {
    cat(e$message)
  }
)
```

Instead, we'll need partial function evaluation, which will create a function with one a
ellipsis (`...`) argument that will fill in the values you specified: 

```{r}
# Create my partially evaluated function
sim_twin_tree_function <- pryr::partial(
  sim_bd_twin_tree, 
  seed = 314, 
  method = "random_tree", 
  n_replicates = 1
)
# Create twinning_params with my function 
twinning_params <- create_twinning_params(
  sim_twin_tree_function = sim_twin_tree_function
)
# Create a twin tree using my function
twin_tree <- pirouette::create_twin_tree(
  phylogeny = true_phylogeny,
  twinning_params = twinning_params
)
# Show the twin tree
ape::plot.phylo(twin_tree)
```

Using partial function evaluation you can now plug in any function to create
a twin tree.

## 2. create a true alignment

We start from a simple true tree:

```{r}
true_phylogeny <- ape::read.tree(text = "((A:1, B:1):1, C:2);")
ape::plot.phylo(true_phylogeny)
```

We want to create a true alignment from it. There is a function that does
so:

```{r}
true_alignment <- sim_true_alignment(true_phylogeny)
ape::image.DNAbin(true_alignment)
```

The `sim_true_alignment`, however, creates the true alignment in the way specified
in its other function argument: the `alignment_params`.

```{r}
alignment_params <- create_alignment_params()
```

Calling `sim_true_alignment` results in the same true alignment:

```{r}
true_alignment <- sim_true_alignment(
  true_phylogeny = true_phylogeny,
  alignment_params = alignment_params
)
ape::image.DNAbin(true_alignment)
```

How to specify *how* the true alignment is created? Use 
the `alignment_params$sim_true_alignment_function`:

```{r}
alignment_params <- create_alignment_params(
  sim_true_alignment_function = get_default_sim_true_alignment_function()
)
true_alignment <- sim_true_alignment(
  true_phylogeny = true_phylogeny,
  alignment_params = alignment_params
)
ape::image.DNAbin(true_alignment)
```

Due to this architecture, `alignment_params$sim_true_alignment_function` must
be a function that takes *one* argument, which is the true phylogeny.

Here is an example that creates a true alignment by simulating 
a random alignment with the same number of taxa:

```{r}
# My function
my_function <- function(true_phylogeny) {
  sequences <- list()
  for (i in seq_len(ape::Ntip(true_phylogeny))) {
    sequences[[i]] <- rep(sample(c("a", "c", "g", "t"), size = 1), 1000)
  }
  ape::as.DNAbin(sequences)
}
# Putting my function in the alignment_params
alignment_params <- create_alignment_params(
  sim_true_alignment_function = my_function
)
# Simulate the true alignment using my function
true_alignment <- pirouette::sim_true_alignment(
  true_phylogeny = true_phylogeny,
  alignment_params = alignment_params
)
# Show the true alignment
ape::image.DNAbin(true_alignment)
```

So, `alignment_params$sim_true_alignment_function` must
be a function that takes *one* argument, which is the true phylogeny. What
if we want a function that indeed has one argument for the true phylogeny,
but also some additional ones?

An example is `sim_true_alignment_with_standard_site_model`, which uses a 
root sequence of `acgt`, a mutation rate of 0.1 and a JC69 site model:

```{r}
head(sim_true_alignment_with_standard_site_model)
```

To change the default arguments, we'll need partial function evaluation, 
which will create a function with one a
ellipsis (`...`) argument that will fill in the values you specified: 

```{r}
sim_true_alignment_function <- pryr::partial(
  sim_true_alignment_with_standard_site_model, 
  root_sequence = "aaaaaaaa", 
  mutation_rate = 0.5,                     
  site_model = beautier::create_hky_site_model()
)
head(sim_true_alignment_function)
```

```{r}
alignment_params <- create_alignment_params(
  sim_true_alignment_function = sim_true_alignment_function
)
true_alignment <- sim_true_alignment(
  true_phylogeny = true_phylogeny,
  alignment_params = alignment_params
)
ape::image.DNAbin(true_alignment)
```

Instead of doing the partial function evaluation yourself, `pirouette`
supplies this function as `get_sim_true_alignment_with_standard_site_model_function`:

```{r}
alignment_params <- create_alignment_params(
  sim_true_alignment_function = 
    get_sim_true_alignment_with_standard_site_model_function(
    root_sequence = "aaaaaaaa", 
    mutation_rate = 0.5,                     
    site_model = beautier::create_hky_site_model()
  )
)
true_alignment <- sim_true_alignment(
  true_phylogeny = true_phylogeny,
  alignment_params = alignment_params
)
ape::image.DNAbin(true_alignment)
```

Or to use the linked node substitution model, use
`get_sim_true_alignment_with_linked_node_sub_site_model_function`:

```{r}
alignment_params <- create_alignment_params(
  sim_true_alignment_function = 
    get_sim_true_alignment_with_linked_node_sub_site_model_function(
    subst_matrix = NULL,
    branch_mutation_rate = 0.1,
    node_mutation_rate = 0.2,
    root_sequence = "cccccccc"
  )
)
true_alignment <- sim_true_alignment(
  true_phylogeny = true_phylogeny,
  alignment_params = alignment_params
)
ape::image.DNAbin(true_alignment)
```

Or to use the unlinked node substitution model, use
`get_sim_true_alignment_with_unlinked_node_sub_site_model_function`:

```{r}
alignment_params <- create_alignment_params(
  sim_true_alignment_function = 
    get_sim_true_alignment_with_unlinked_node_sub_site_model_function(
    branch_mutation_rate = 1.0,
    node_mutation_rate = 2.0,
    node_time = 0.1,
    root_sequence = "aaaaaaaa"
  )
)
true_alignment <- sim_true_alignment(
  true_phylogeny = true_phylogeny,
  alignment_params = alignment_params
)
ape::image.DNAbin(true_alignment)
```


Using partial function evaluation you can now plug in any function to create
a true alignment.



## 3. create a twin alignment

To create a twin alignment, we first need a twin tree and a true alignent to 
work on.

As the twin tree, we'll use this:

```{r}
twin_phylogeny <- ape::read.tree(text = "((A:2, B:2):1, C:3);")
ape::plot.phylo(twin_phylogeny, main = "Twin phylogeny")
```
As a true alignment, we'll use something simple:

```{r}
x <- list()
x[[1]] <- rep("c", 4)
x[[2]] <- rep("g", 4)
x[[3]] <- rep("t", 4)
names(x) <- c("A", "B", "C")
true_alignment <- ape::as.DNAbin(x)
ape::image.DNAbin(true_alignment, main = "True alignment", legend = FALSE)
testit::assert(get_alignment_n_taxa(true_alignment) == 3)
```

Creating a twin alignment:

```{r}
twin_alignment <- create_twin_alignment(
  twin_phylogeny = twin_phylogeny,
  true_alignment = true_alignment,
  alignment_params = create_test_alignment_params(),
  twinning_params = create_twinning_params()
)
ape::image.DNAbin(twin_alignment, main = "Twin alignment")
```

Creating a twin alignment from this:

```{r}
twin_alignment <- create_twin_alignment(
  twin_phylogeny = twin_phylogeny,
  true_alignment = true_alignment,
  alignment_params = create_test_alignment_params(),
  twinning_params = create_twinning_params()
)
ape::image.DNAbin(twin_alignment)
```

Now we'll create our function we'll plug in:

```{r}
create_rnd_alignment <- function(twin_phylogeny, true_alignment) {
  n_taxa <- ape::Ntip(twin_phylogeny)
  n_nucleotides <- pirouette::get_alignment_sequence_length(true_alignment)
  x <- list()
  for (i in seq_len(n_taxa)) {
    x[[i]] <- sample(x = c("a", "c", "g", "t"), size = n_nucleotides, replace = TRUE)
  }
  ape::as.DNAbin(x)
}
```

Plugging it in:

```{r}
twin_alignment <- create_twin_alignment(
  twin_phylogeny = twin_phylogeny,
  true_alignment = true_alignment,
  alignment_params = create_test_alignment_params(),
  twinning_params = create_twinning_params(
    sim_twin_alignment_function = create_rnd_alignment
  ),
  newskool = TRUE
)
ape::image.DNAbin(twin_alignment)
```


Now we want to plug in:

```{r}
head(create_twin_alignment_with_standard_site_model)
```

Here comes partial function evaluation to the rescue:

```{r}
root_sequence <- create_blocked_dna(get_alignment_sequence_length(true_alignment))  
sim_twin_align_function <- pryr::partial(
  create_twin_alignment_with_standard_site_model,
  root_sequence = root_sequence,
  mutation_rate = 0.1
)
head(sim_twin_align_function)
check_sim_twin_alignment_function(sim_twin_align_function)
```



Plugging it in:

```{r}
twin_alignment <- create_twin_alignment(
  twin_phylogeny = twin_phylogeny,
  true_alignment = true_alignment,
  alignment_params = create_test_alignment_params(),
  twinning_params = create_twinning_params(
    sim_twin_alignment_function = sim_twin_align_function
  ),
  newskool = TRUE
)
ape::image.DNAbin(twin_alignment)
```
