---
title: "Tutorial"
author: "Richel J.C. Bilderbeek"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates the `pirouette` package, 
with a focus on the research questions that `pirouette` can answer.

# Main question

The main question that `pirouette` answers is:

> What is the error BEAST2 makes in inferring a given phylogeny?

The same question, but from the point of view of a theoretician:

> Would we know the true phylogeny in nature,
> how well would we be able to infer it 
> from a DNA alignment?

# Inferring a simplest world

Let's assume we know speciation (we don't) and we assume that speciation
is as simple as possible (it is not). 

In that simplest world:

 * Speciation follows a (constant-rate) birth-death process
 * DNA sequences of all species have the same mutation rate
 * In DNA mutations, mutation rates between all nucleotides are equal

We can now try to answer `pirouette`'s main question: 

> Would we know the true phylogeny in nature,
> how well would we be able to infer it 
> from a DNA alignment?

Here we go, the true phylogeny of the world:

```{r fig.width=7, fig.height=7}
set.seed(42)
phylogeny <- becosys::bd_sim(
  bd_params = becosys::create_bd_params(spec_rate = 1.0,  ext_rate = 0.1),
  crown_age = 2.0,
  conditioned_on = "non_extinction"
)$reconstructed_tree
ape::plot.phylo(main = "The True Phylogeny", phylogeny)
```
 
This phylogeny cannot be measured in nature directly. Instead, we can measure 
(and align) the DNA sequences of our species. 

`pirouette` needs an approximate idea of how an alignment in nature behaves.

What we do know (or: assumed in our simples world) 
is (1) that the DNA sequences of all species have the same 
mutation rate, and (2) In DNA mutations, mutation rates between all nucleotides 
are equal. This amounts to (1) the clock model is a strict clock model, 
and (2) the site model is a Jukes-Cantor model.

What we need to have approximately right is a mutation rate, which must not
be too low, nor too high.
A mutation rate that is too low (e.g. zero) would result in an uninformative 
alignment (e.g. all species would have the same sequence). 
A mutation rate that is too high, results in uninformative alignments as well,
consisting of random noise.

What we do for convenience is specify an ancestral DNA sequence
that is simple and easy to interpret visually, as well 
as an RNG seed:

```{r}
alignment_params = pirouette::create_alignment_params(
  site_model = beautier::create_jc69_site_model(),
  clock_model = beautier::create_strict_clock_model(),
  mutation_rate = 0.3,
  root_sequence = "aaaaaaaaccccccccggggggggtttttttt", 
  rng_seed = 314
)
```

From here, we can already predict the DNA alignment `pirouette` will generate:

```{r fig.width=7, fig.height=7}
alignment <- pirouette::sim_alignment(
  phylogeny = phylogeny, 
  alignment_params = alignment_params
) 
ape::image.DNAbin(
  alignment,
  main = "A Possibly-True Alignment",
  legend = FALSE
)
```

Back to our main question:

> Would we know the true phylogeny in nature,
> how well would we be able to infer it 
> from a DNA alignment?

To infer a phylogeny from an alignment, we need:

 * a way to select inference parameters
 * a general setup for our inference

We select our inference parameters based on our knowledge of the
generative model. The generative model is the model we followed in simulating
the DNA alignment. We do need to specify a tree prior, we we know,
due to our knowledge of The Simplest World.

```{r}
gen_model_select_param <- pirouette::create_gen_model_select_param(
  alignment_params = alignment_params,
  tree_prior = beautier::create_bd_tree_prior()
)
```

The general setup for our inference can be tuned. For example,
to have a shorter MCMC run and a specific RNG seed:

```{r}
inference_params <- pirouette::create_inference_params(
  mcmc = beautier::create_mcmc(chain_length = 2000, store_every = 1000),
  rng_seed = 271
)
```

Now we have all in place to do a `pirouette` run:

```{r}
df <- pirouette::pir_run(
  phylogeny = phylogeny,
  alignment_params = alignment_params,
  model_select_params = gen_model_select_param,
  inference_params = inference_params
)
knitr::kable(df)
```

Or in a violin plot:

```{r fig.width=7, fig.height=7}
pirouette::pir_plot(df)
```

# Doing inference with the best tree prior

From the Simplest World scenarion, we drop one piece of knowledge:
we assumre we do not know the tree prior anymore. We do still know
The True Phylogeny, but do not know the speciation model that generated 
it (even though we generated the phylogeny using a birth-death model).

> Would we know the true phylogeny in nature,
> how well would we be able to infer it 
> from a DNA alignment?

So, again, here is the true phylogeny:

```{r fig.width=7, fig.height=7}
ape::plot.phylo(main = "The True Phylogeny", phylogeny)
```

Also, the alignment it generated in the Simplest World
as well is still equally valid:

```{r fig.width=7, fig.height=7}
ape::image.DNAbin(
  alignment,
  main = "A Possibly-True Alignment",
  legend = FALSE
)
```

Because we do not know the true tree prior, we instruct `pirouette`
to use the best model to do inference. This best model is determined
on the evidence (also: marginal likelihood) for a model in generating
the alignment. We can specify the set of models `pirouette` can pick from.
In this case, we let `pirouette` pick from all (only two!) model combinations
of one site model (JC69), one clock model (strict) and two tree priors.
We choose `pirouette` to prefer a quick calculation over a
correct estimation by setting `epsilon` to a high value:

```{r}
best_model_select_param <- pirouette::create_best_model_select_param(
  site_models = list(beautier::create_jc69_site_model()),
  clock_models = list(beautier::create_strict_clock_model()),
  tree_priors = list(
    beautier::create_yule_tree_prior(),
    beautier::create_bd_tree_prior()  
  ),
  epsilon = 100.0
)
```

Now we again have all in place to do a `pirouette` run:

```{r fig.width=7, fig.height=7}
if (rappdirs::app_dir()$os != "win") {
  df <- pirouette::pir_run(
    phylogeny = phylogeny,
    alignment_params = alignment_params,
    model_select_params = best_model_select_param,
    inference_params = inference_params
  )
  knitr::kable(df)
  pirouette::pir_plot(df)
}
```

We can also specify our complete knowledge at the same time:

 * the generative model is JC69, strict and BD
 * pick the best inference model

```{r fig.width=7, fig.height=7}
if (rappdirs::app_dir()$os != "win") {
  df <- pirouette::pir_run(
    phylogeny = phylogeny,
    alignment_params = alignment_params,
    model_select_params = list(gen_model_select_param, best_model_select_param),
    inference_params = inference_params
  )
  knitr::kable(df)
  pirouette::pir_plot(df)
}
```
